### 2017-04-12 src/stats6.tex
Added "Analysis of Variance" to SS6.

        \end{example}
\newpage
\section{Analysis of Variance}

>>>>

\newpage
\section{Analysis of Variance}
ANOVA tests are an extension of F-Tests for tests with more than 2 populations. When dealing with anova tests their are two new terms used:
\paragraph{Factor} A characteristic under consideration, thought to influence the measured observations.
\paragraph{Level} A value of the factor.\\
\\
Consider the following test:
A car magazine wishes to compare the average petrol consumption of three simular models of car and has availible six vehicles of each model. \\
\\
This test has 1 factor (model of car) at 3 levels (3 diffrent models).

    \newpage
    \subsection{One-way ANOVA}
        A one-way ANOVA test is used when their is only 1 factor. The follwing notation and formule are used in the calculation of a one-way ANOVA test:
        \bgroup
        \def\arraystretch{2}
        \begin{center}
        \begin{tabular}{l|c|c}
        Description & Symbol & Value \\
        \hline
        Number of samples/levels & k & \\
        Number of observations in $i^{th}$ sample & $n_i$ & $i = 1, 2, ..., k$ \\
        Total number of observations & $n$ & $\displaystyle\sum_i{n_i}$ \\
        Observation $j$ in $i^{th}$ sample & $x_{ij}$ & $j = 1, 2, ..., n_i$ \\
        Sum of $n_i$ observations in the $i^{th}$ sample & $T_i$ & $\displaystyle\sum_j{x_{ij}}$ \\
        Sum of all n observations & $T$ & $\displaystyle\sum_i{T_i}$ \\
        \hline
        Total sum of squares & $SS_T$ & $\displaystyle\sum_i{\displaystyle\sum_j{x_{ij}^2}} - \frac{T^2}{n}$ \\
        Between samples sum of squares & $SS_B$ & $\displaystyle\sum_i{\frac{T_i}{n_i}} - \frac{T^2}{n}$ \\
        Within samples sum of squares & $SS_W$ & $SS_T - SS_B$ \\
        \hline
        Between samples mean squares & $MS_B$ & $\displaystyle\frac{SS_B}{k - 1}$ \\
        Within samples mean squares& $MS_W$ & $\displaystyle\frac{SS_W}{n - k}$ \\
        Total mean squares & $MS_T$ & $\displaystyle\frac{SS_T}{n - 1}$ \\
        \end{tabular}
        \end{center}
        \egroup

        \noindent
        Its convient to lay out an ANOVA test in a table such as:
        \bgroup
        \def\arraystretch{2}
        \begin{center}
        \begin{tabular}{l|c|c|c|c}
        Source of variation & Sum of squares    & Degrees of freedom    & Mean square   & F-ratio               \\
        \hline
        Between samples     & $SS_B$            & $k-1$                 & $MS_B$        & $\dfrac{MS_B}{MS_W}$  \\
        Within samples      & $SS_W$            & $n-k$                 & $MS_W$        &                       \\
        \hline
        Total               & $SS_T$            & $n-1$                 &               &                       \\
        \end{tabular}
        \end{center}
        \egroup
        The F-ratio forms the test statistic and the F-distribution is used to find the critical value. Note that $MS_B$ will always be larger than $MS_W$ so we always select the upper tail as the critical value.

        \begin{example}
        {   
            In a comparision of the cleaning action of four detergents, pieces of cloth were measured for whiteness after being washed by a particular detergent. The results were as follows:
            \begin{center}
            \begin{tabular}{c|c|c|c}
            A  & B  & C  & D  \\
            \hline
            77 & 74 & 73 & 76 \\
            81 & 66 & 78 & 85 \\
            61 & 58 & 57 & 77 \\
            76 &    & 69 & 64 \\
            69 &    & 63 &    \\
            \end{tabular}
            \end{center}
            Assuming the whiteness readings are normally distributed, with a common variance, test hypothesis that their is no diffrence between the four brands whiteness using a 5\% significance level.
        }

        \begin{step}{Hypothesis}
        $H_0$: no diffrence in the mean whiteness\\
        $H_1$: a diffrence in the mean whiteness\\
        \end{step}

        \begin{step}{Work out the sum of squares}
        \begin{center}
        \begin{tabular}{l|c|c|c|c|c}
                & A     & B     & C     & D     & Total \\
        \hline
        $n_i$   & 5     & 3     & 5     & 4     & 17    \\
        $T_i$   & 364   & 198   & 340   & 302   & 1204  \\
        \end{tabular}
        \end{center}

        \begin{align*}
        SS_T &= 77^2 + 81^2 + 61^2 + ... + 64^2 - \frac{1204^2}{17}\\
        &= 86362 - 85271.529...\\
        &= 1090.47\\
        \\
        SS_B &= (\frac{364^2}{5} + \frac{198^2}{3} + \frac{340^2}{5} + \frac{302^2}{4}) - \frac{1204^2}{17}\\
        &= 85488.2 - 85271.529...\\
        &= 216.67\\
        \\
        SS_W &= 1090.47 - 216.67\\
        &= 873.80
        \end{align*}
        \end{step}

        \begin{step}{Complete the ANOVA table}
        \begin{center}
        \begin{tabular}{l|c|c|c|c}
        Source of variation & Sum of squares    & Degrees of freedom    & Mean square   & F-ratio   \\
        \hline
        Between samples     & $216.67$          & $3$                   & $72.22$       & $1.07$    \\
        Within samples      & $873.80$          & $13$                  & $67.22$       &           \\
        \hline
        Total               & $1090.47$         & $16$                  &               &           \\
        \end{tabular}
        \end{center}
        \end{step}

        \begin{step}{Critical value}
        $$
        F_{(3, 13)}(0.95) = 3.411 
        $$
        \end{step}

        \begin{step}{Conclusion}
        $1.07 < 3.411$ so accept $H_0$, their is no evidence at the 5\% significance level to suggest that their is any diffrence between the four brands.
        \end{step}
        \end{example}

    \newpage
    \subsection{Two-way ANOVA}
        A two-way ANOVA is an extension to the one-way ANOVA test that takes into account a second factor. The second factor is the blocking factor (since this test requires a radomised block design) since it places the samples into blocks.
        $$
        x_{ij} - \mu_{ij} = \epsilon_{ij} \sim N(0, \sigma^2)
        $$
        where
        $$
        \mu_{ij} = \mu + \alpha_i + \beta_j
        $$
        where $\mu$ is the overall mean, $\alpha_i$ is the mean effect of the $i^{th}$ level of the row factor relative to $\mu$, $\beta_j$ is the mean effect of the $j^{th}$ level of the column factor relative to $\mu$ and $\epsilon_{ij}$ is the inherent random variation.\\
        \\
        The follwing notation and formule are used in the calculation of a two-way ANOVA test:
        \bgroup
        \def\arraystretch{2}
        \begin{center}
        \begin{tabular}{l|c|c}
        Description & Symbol & Value \\
        \hline
        Number of levels of row factor & $m$ & \\
        Number of levels of column factor & $n$ & \\
        Total number of observations & $mn$ & \\
        Observation in $ij^{th}$ cell & $x_{ij}$ & $i = 1, 2, ..., m$ \\
        & & $j = 1, 2, ..., n$  \\
        Sum of $n$ observations in the $i^{th}$ row & $R_i$ & $\sum_j{x_{ij}}$ \\
        Sum of $m$ observations in the $j^{th}$ row & $C_j$ & $\sum_i{x_{ij}}$ \\
        Sum of all $mn$ observations & $T$ & $\sum_i{\sum_j{x_{ij}}} = \sum_i{R_i} = \sum_j{R_j}$ \\
        \hline
        Total sum of squares & $SS_T$ & $\displaystyle\sum_i{\displaystyle\sum_j{x_{ij}^2}} - \frac{T^2}{mn}$ \\
        Between rows sum of squares & $SS_R$ & $\displaystyle\sum_i{\frac{R_i^2}{n}} - \frac{T^2}{mn}$ \\
        Between column sum of squares & $SS_C$ & $\displaystyle\sum_i{\frac{C_j^2}{m}} - \frac{T^2}{mn}$ \\
        Error (residual) sum of squares & $SS_E$ & $SS_T - SS_R - SS_C$ \\
        \hline
        Between row mean squares & $MS_R$ & $\displaystyle\frac{SS_R}{n - 1}$ \\
        Between column mean squares & $MS_C$ & $\displaystyle\frac{SS_C}{m - 1}$ \\
        Error (residual) mean squares & $MS_E$ & $\displaystyle\frac{SS_E}{mn - 1}$ \\
        \end{tabular}
        \end{center}
        \egroup
        \noindent

        Its convient to lay out a two-way ANOVA test in a table such as:
        \bgroup
        \def\arraystretch{2}
        \begin{center}
        \begin{tabular}{l|c|c|c|c}
        Source of variation & Sum of squares    & Degrees of freedom    & Mean square   & F-ratio               \\
        \hline
        Between rows        & $SS_R$            & $m-1$                 & $MS_R$        & $\dfrac{MS_R}{MS_E}$  \\
        Between columns     & $SS_C$            & $n-1$                 & $MS_C$        & $\dfrac{MS_C}{MS_E}$  \\
        Error (residual)    & $SS_E$            & $(m-1)(n-1)$          & $MS_E$        &                       \\
        \hline
        Total               & $SS_T$            & $mn-1$                &               &                       \\
        \end{tabular}
        \end{center}
        \egroup

        \begin{example}
        {
            Test at a 1\% significance level, that their is no diffrence between the four compilers and was the use of the program as a blocking factor worthwhile.

            \begin{center}
            \begin{tabular}{c|c|c|c|c}
            & \multicolumn{4}{c}{Compiler} \\
            Program & 1         & 2         & 3         & 4         \\
            \hline
            A       & $29.21$   & $28.25$   & $28.20$   & $28.62$   \\
            B       & $26.18$   & $26.02$   & $26.22$   & $25.56$   \\
            C       & $30.91$   & $30.18$   & $30.52$   & $30.09$   \\
            D       & $25.14$   & $25.26$   & $25.20$   & $25.02$   \\
            E       & $26.16$   & $25.14$   & $25.26$   & $25.46$   \\
            \end{tabular}
            \end{center}
        }

        \begin{step}{Hypothesis}
        $H_0$: No effect on compilation times due to compilers\\
        $H_1$: An effect on compilation times due to compilers\\
        \end{step}

        \begin{step}{Find the row and column totals}
        \begin{center}
        \begin{tabular}{c|cccc|c|c}
                            & \multicolumn{4}{c}{Compiler}                  & Row totals    &           \\
        Program             & 1         & 2         & 3         & 4         & $T_i$         & $n_i$     \\
        \hline          
        A                   & $29.21$   & $28.25$   & $28.20$   & $28.62$   & $114.28$      & 4         \\
        B                   & $26.18$   & $26.02$   & $26.22$   & $25.56$   & $103.98$      & 4         \\
        C                   & $30.91$   & $30.18$   & $30.52$   & $30.09$   & $121.70$      & 4         \\
        D                   & $25.14$   & $25.26$   & $25.20$   & $25.02$   & $100.62$      & 4         \\
        E                   & $26.16$   & $25.14$   & $25.26$   & $25.46$   & $102.02$      & 4         \\
        \hline
        Column totals $T_j$ & $137.60$  & $134.85$  & $135.40$  & $134.75$  & $T = 542.60$  &           \\
        \hline
        $n_j$               & $5$       & $5$       & $5$       & $5$       &               & $mn = 20$ \\
        \end{tabular}
        \end{center}
        \end{step}

        \begin{step}{Find the sum of squares}
        \begin{align*}
        SS_T &= (29.21^2 + 28.25^2 + ... + 25.46^2) - \frac{542.60^2}{20} \\
        &= 14805.7768 - 14720.738\\
        &= 85.0388\\
        \\
        SS_R &= \left(\frac{114.28^2}{4} + \frac{103.98^2}{4} + \frac{121.70^2}{4} + \frac{100.62^2}{4} + \frac{102.02^2}{4}\right) - \frac{542.60^2}{20}\\
        &= 83.0404\\
        \\
        SS_C &= \left(\frac{137.60^2}{5} + \frac{134.85^2}{5} + \frac{135.40^2}{5} + \frac{134.75^2}{5}\right) - \frac{542.60^2}{20}\\
        &= 1.063\\
        \\
        SS_E &= 85.0388 - 83.0404 - 1.063\\
        &= 0.9354
        \end{align*}
        \end{step}

        \begin{step}{Complete the ANOVA table}
        \begin{center}
        \begin{tabular}{l|c|c|c|c}
        Source of variation & Sum of squares    & Degrees of freedom    & Mean square   & F-ratio       \\
        \hline
        Between rows        & $83.0404$         & $4$                   & $20.7601$     & $266.33$      \\
        Between columns     & $1.063$           & $3$                   & $0.3543$      & $4.55$        \\
        Error (residual)    & $0.9354$          & $12$                  & $0.07795$     &               \\
        \hline
        Total               & $85.0388$         & $11$                  &               &               \\
        \end{tabular}
        \end{center}
        \end{step}

        \begin{step}{Critical value}
        $$
        F_{(3, 12)}(0.99) = 5.953
        $$
        \end{step}

        \begin{step}{Conclusion}
        $4.55 < 5.953$ so accept $H_0$, thus their is no evidence at the 1\% significance level to suggest a diffrence in compilation times between the four compilers.
        \end{step}

        \begin{step}{Was the blocking factor worthwhile?}
        \begin{itemize}
        \item The variation between programs accounted for $\dfrac{83.0404}{85.0388} \times 100 = 97.65\%$ of the variation (which most would have gone into $SS_E$ had the blocking factor not been used).
        \item $F_R = 266.33$ which would have been significant at any level.
        \end{itemize}
        \end{step}

    \newpage
    \subsection{Latin square designs}
        Consider a randomised block design in which:
        $$ \text{number of blocks} = \text{number of treatments} = n $$
        This is equivilent to a two-way ANOVA in which:
        $$ \text{number of rows} = \text{number of columns} = n $$
        So the total number of observations is $n^2$.\\
        \\
        It is possible to intoduce a second blocking factor (with $n$ levels) without increasing the number of observations to $n^3$ with the following model:
        $$
        x_{ij(k)} - \mu_{ij(k)} = \epsilon_{ij(k)} \sim N(0, \sigma^2)
        $$
        where 
        \begin{itemize}
        \item $x_{ij(k)}$ is the observation in the $i^{th}$ row, $j^{th}$ column with the letter $k$
        \item $\mu$ is the overall mean
        \item $\alpha_i$ is the mean effect of the $i^{th}$ level of row factor relative to $\mu$
        \item $\beta_j$ is the mean effect of the $j^{th}$ level of column factor relative to $\mu$
        \item $\gamma_k$ is the mean effect of the $k^{th}$ level of letter factor relative to $\mu$
        \item $\epsilon_{ij(k)}$ is the inherent random variation 
        \end{itemize}

        \begin{example}
        {
            In a comparision of three types of fuels (A, B and C), three makes of cars (1, 2 and 3), were used. Each car was driven at a speed of $50 km h^-1$, $75 km h^-1$ and $100 km h^-1$ around a track. The distrance travelled from depleting a full tank was recorded.

            \begin{center}
            \begin{tabular}{c|c|c|c}
            Car & Speed & Fuel  & Distance  \\
            1   & 50    & B     & 347.8     \\
            1   & 75    & C     & 331.4     \\
            1   & 100   & A     & 275.8     \\
            2   & 50    & A     & 365.4     \\
            2   & 100   & B     & 339.8     \\
            2   & 150   & C     & 298.9     \\
            3   & 50    & C     & 318.9     \\
            3   & 100   & A     & 302.2     \\
            3   & 150   & B     & 271.8     \\
            \end{tabular}
            \end{center}

            Investigate the diffrence between the three petrols.
        }

        \begin{step}{Rearrange the data into a latin square}
        \begin{center}
        \begin{tabular}{c|ccc|c|c}
                        & \multicolumn{3}{c}{Speed}         &           &           \\
        Car             & 50        & 75        & 100       & Row Total & $n_i$     \\
        \hline
        1               & B $347.8$ & C $331.4$ & A $275.8$ & $955.0$       & 3     \\
        2               & A $365.4$ & B $339.8$ & C $298.9$ & $1004.1$      & 3     \\
        3               & C $318.9$ & A $302.2$ & B $271.8$ & $892.9$       & 3     \\
        \hline
        Column total    & $1032.1$  & $973.4$   & $846.5$   & $T = 2852.0$  &       \\
        \hline
        $n_j$           & 3         & 3         & 3         &               & n = 9 \\
        \end{tabular}
        \end{center}

        \begin{align*}
        L_A &= 943.4\\
        L_B &= 959.4\\
        L_C &= 949.2
        \end{align*}
        \end{step}

        \begin{step}{Find the sum of squares}
        \begin{align*}
        SS_T &= (347.8^2 + 331.4^2 + ... + 271.8^2) - \frac{2852.0^2}{9}\\
        &= 912076.14 - 903767.11\\
        &= 8309.03\\
        \\
        SS_R &= (\frac{955.0^2}{3} + \frac{1004.1^2}{3} + \frac{892.9^2}{3}) - 903767.11\\
        &= 2070.30\\
        \\
        SS_C &= (\frac{1032.1^2}{3} + \frac{973.4^2}{3} + \frac{846.5^2}{3}) - 903767.11\\
        &= 5999.63\\
        \\
        SS_L &= (\frac{943.4^2}{3} + \frac{959.4^2}{3} + \frac{949.2^2}{3}) - 903767.11\\
        &= 43.74\\
        \\
        SS_E &= 8309.03 - 2070.30 - 5999.63 - 43.74
        &= 195.36
        \end{align*}
        \end{step}

        \begin{step}{Contruct ANOVA table}
        \begin{center}
        \begin{tabular}{l|c|c|c|c}
        Source of variation & Sum of squares    & Degrees of freedom    & Mean square   & F-ratio       \\
        \hline
        Between cars        & $2070.30$         & $2$                   & $1035.15$     & $10.60$       \\
        Between speeds      & $5999.63$         & $2$                   & $2999.81$     & $30.71$       \\
        Between brands      & $43.74$           & $2$                   & $21.87$       & $0.22$        \\
        Residual            & $0.9354$          & $2$                   & $97.68$       &               \\
        \hline
        Total               & $8309.03$         & $8$                   &               &               \\
        \end{tabular}
        \end{center}
        \end{step}

        \begin{step}{Critical value}
        $$
        F_{(2, 2)}(0.95) = 19.00
        $$
        \end{step}

        \begin{step}{Conclusion}
        \begin{itemize}
        \item $0.22 < 19.00$ so their is no evidence that fuel consumption is dependent on the brand of fuel
        \item $30.71 > 19.00$ so the speed of cars proved to be signigicant
        \item $10.60 < 19.00$ so the car was not suggnificant, however it did reduce the residual sum of squares 
        \end{itemize}
        \end{step}
